---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

* Read the activity data
* account for the intervals / plot jump issues by converting interval to actual minutes since start of day (e.g. 115 -> 75 minutes)
* Create a filtered dataset, omitting the NA values
* Calculate the Daily Total Steps 
* Calculate the Total and Mean Steps for each interval across days
* Setup some convenience variables for the x-axis in plots

```{r,results=FALSE,message=FALSE}
require(dplyr)
require(stringr)
activity.file <- unzip("activity.zip")
activity.data <- read.csv(activity.file)
interval.labels <- unique(paste( substr(str_pad(activity.data$interval,4,pad="0"),1,2),":",substr(str_pad(activity.data$interval,4,pad="0"),3,4), sep=""))
activity.data <- mutate(activity.data, interval = floor(interval/100)*60 + interval %% 60)
activity.data.filtered <- na.omit(activity.data)
daily.steps <- activity.data.filtered  %>% group_by(date) %>% summarise(total=sum(steps)) 
interval.total.steps <- activity.data.filtered %>% group_by(interval) %>% summarise(total=sum(steps))
interval.mean.steps <- activity.data.filtered %>% group_by(interval) %>% summarise(stepmean=mean(steps))

plot.at <- c(1,180,360,540,720,900,1080,1260,1435)
plot.xlabels <- c(interval.labels[1],interval.labels[37],interval.labels[73],interval.labels[109],interval.labels[145],interval.labels[181],interval.labels[217],interval.labels[253],interval.labels[288])


```


## What is mean total number of steps taken per day?

* Calculate the Mean and Median for the Total Daily Steps
* Plot the Histogram of Total Daily Steps

```{r}
mean.daily.steps <- mean(daily.steps$total)
median.daily.steps <-median(daily.steps$total)
par(bg = "lightskyblue1")

hist(daily.steps$total, xlab="Total Daily Steps", main= "Histogram of Total Daily Steps")

```

+ **Mean Total Daily Steps   :  _`r format(mean.daily.steps,digits=7)`_**
+ **Median Total Daily Steps : _`r format(median.daily.steps,digits=7)`_**


## What is the average daily activity pattern?

* Determine the max step interval
* Create the plot of Average Steps Across Days by Interval

```{r}
max.step.interval <- interval.mean.steps$interval[which.max(interval.mean.steps$stepmean)]
par(bg = "lightskyblue1")
plot(interval.mean.steps ,type="l", xlab="Time Interval", ylab="Average Number of Steps", main = "Average Steps Across Days By Interval",xaxt="n")
axis(1, at=plot.at ,labels= plot.xlabels)
abline(v=max.step.interval, col="yellow" , lwd=1, lty =2)
text(paste(floor(max.step.interval/60),":",max.step.interval%%60,sep=""), x= max.step.interval, y = max(interval.mean.steps$stepmean),col="blue")
```


+ **The 5 minute interval across all days in the dataset containing the maximum number of total steps occurs at  _`r paste(floor(max.step.interval/60),":",max.step.interval%%60,sep="") `_**



## Imputing missing values

* Determine the count of missing values

```{r}
na.steps <- sum(is.na(activity.data$steps))
```

1. **The  number of rows in the dataset missing  steps is   _`r na.steps `_  (out of _`r nrow(activity.data)`)_**


* Compute the total steps for a given interval across all days 
* Determine the average steps for a given interval (Total from previous steps divided by the number of days)
* For a day/interval combination that is missing data, use the interval step average computed from the previous step as the imputed value

```{r}
activity.data.imputed <- activity.data
for(i in 1:nrow(activity.data)) {
    if(is.na(activity.data[i,1]) == TRUE) {
      activity.data.imputed[i,1] <- round(interval.total.steps$total[which(interval.total.steps$interval ==  activity.data[i,3])]/length(unique(activity.data.filtered$date)))
    }
}
daily.steps.imputed <- activity.data.imputed  %>% group_by(date) %>% summarise(meansteps=sum(steps)) 
mean.daily.steps.imupted <- mean(daily.steps.imputed$meansteps)
median.daily.steps.imputed <-median(daily.steps.imputed$meansteps)


par(bg = "lightskyblue1")
hist(daily.steps.imputed$meansteps, xlab="Total Daily Steps(imputed missing values)", main= "Histogram of Total Daily Steps")

```


+ **Mean Total Daily Steps  (with imputed data) :  _`r format(mean.daily.steps,digits=7)`_**
+ **Median Total Daily Steps (with imputeddata) : _`r format(median.daily.steps,digits=7)`_**


## Are there differences in activity patterns between weekdays and weekends?

* Create a factor column to detemine Weekend/Weekday
* Compute the Average Steps across weekdays, weekends by interval
* Generate the required panel xyplot 


```{r message=FALSE}

activity.data.imputed<- 
  mutate(activity.data.imputed, daytype = 
           factor(ifelse( weekdays(as.Date(date,"%Y-%m-%d")) %in% c("Saturday","Sunday"),"Weekend","Weekday")))

interval.total.steps.imputed <- activity.data.imputed %>% group_by(interval,daytype) %>% summarise(total=mean(steps))
require(lattice)
par(bg = "lightskyblue1")
xyplot(total ~ interval| levels(daytype), 
           data = interval.total.steps.imputed,
           type = "l",
           xlab = "Interval",
           ylab = "Avg. number of steps",
           layout=c(1,2),
           scales =list(x=list(at=plot.at, labels= plot.xlabels)))

```

